---
title: "Final Project – Group Work"
author: "Uriel Ulloa, Jonida Qira, Jordyn Labbe, Ryan Le"
date: "2025-05-05"
output:
  html_document: default
  word_document: default
---

# Analyzing Cars Features to Predict their Prices 
## A Regression Based Approach Using UCI Data 


##  1: Data Loading
#### In this section we load the required data set directly from the source using an URL. 

```{r load the required data}

# Load necessary libraries
library(tidyverse)  
library(readr)
library(ggplot2)

# Define column names
column_names <- c(
  "symboling", "normalized_losses", "make", "fuel_type", "aspiration",
  "num_doors", "body_style", "drive_wheels", "engine_location",
  "wheel_base", "length", "width", "height", "curb_weight",
  "engine_type", "num_cylinders", "engine_size", "fuel_system",
  "bore", "stroke", "compression_ratio", "horsepower", "peak_rpm",
  "city_mpg", "highway_mpg", "price"
)

# Load the dataset using from the UCI URL
url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/autos/imports-85.data"
auto_data <- read_csv(url, col_names = column_names, na = "?")

```


## 2: Data Cleansing
#### The following section explains how we cleaned the data through re-adjusting the data type and dropping any rows with missing data. 

``` {r section2 data cleansing}

# Convert appropriate columns to numeric
numeric_cols <- c("normalized_losses", "bore", "stroke", "horsepower", 
                  "peak_rpm", "price")
auto_data[numeric_cols] <- lapply(auto_data[numeric_cols], as.numeric)

# Drop rows with missing price
auto_data_clean <- auto_data %>% drop_na(price)

# Convert categorical variables to factors
auto_data_clean <- auto_data_clean %>%
  mutate(across(c(make, fuel_type, aspiration, num_doors, body_style,
                  drive_wheels, engine_location, engine_type,
                  num_cylinders, fuel_system), as.factor))

```

## 3: Exploratory Analysis (Response variable)
#### A closer look of response variable distribution is required. We perform this analysis visually by plotting a histogram and boxplot of the Price variable. 

``` {r exploratory analysis}

hist(auto_data_clean$price,
     breaks = 20,               
     col = "skyblue",            
     border = "white",          
     main = "Distribution of Car Prices",
     xlab = "Price (USD)",
     ylab = "Frequency")

boxplot(auto_data_clean$price,
        main = "Boxplot of Vehicle Prices",
        ylab = "Price (USD)",
        col = "skyblue",
        border = "black",
        horizontal = FALSE,
        cex.main = 1.2,
        cex.lab = 1.1)
```

## 4: Variables Selection (Correlation)
#### To start our analysis we decide to explore the relationship between the reponse variable (Price) and each preditors, in order to select which variables to include in the model.

``` {r relationship analysis}
# Select numeric columns
numeric_data <- auto_data_clean %>% select(where(is.numeric))

# Calculate correlations with 'price' only
cor_with_price <- cor(numeric_data, use = "complete.obs")[, "price"]

# View correlations (excluding price-to-price which is always 1)
cor_with_price <- cor_with_price[names(cor_with_price) != "price"]

# Display result
print(cor_with_price)

```

## 5: Variables Selection (Scatter Plot)
#### Following the same line, we decided to graphically validate the linear relationship of the response variable (Price) and each numerical predictor. 


``` {r scatterplot analysis}
# Separate numeric and categorical predictors
numeric_vars <- auto_data_clean %>% 
  select(where(is.numeric)) %>% 
  select(-price) %>% 
  names()

categorical_vars <- auto_data_clean %>% 
  select(where(is.factor)) %>% 
  names()

# Loop over each numeric variable to create scatter plots
for (var in numeric_vars) {
  p <- ggplot(auto_data_clean, aes_string(x = var, y = "price")) +
    geom_point(alpha = 0.6, color = "steelblue") +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    labs(title = paste("Scatter Plot: Price vs", var),
         x = var, y = "Price") +
    theme_minimal()
  print(p)
}

# Ensure categorical variables are treated as factors
auto_data_clean <- auto_data_clean %>%
  mutate(across(where(is.character), as.factor))

# Define quantitative variables to include
quant_vars <- c("wheel_base", "length", "width", "curb_weight", 
                "engine_size", "bore", "horsepower", "city_mpg", 
                "highway_mpg", "peak_rpm")
# Identify qualitative variables (factors)
qual_vars <- auto_data_clean %>% 
  select(where(is.factor)) %>% 
  names()

```


## 6: Categorical Variable Selection
#### Using to many categorical variables, specially the ones with many levels could lead to an overfitting of the model, to avoid this we decide to remove from the model the variables "make" and "fuel system" 

``` {r categorical variables selection}
# Identify qualitative variables (factors) and remove 'make' and 'fuel system'
qual_vars <- auto_data_clean %>% 
  select(where(is.factor)) %>% 
  names() %>%
    setdiff(c("make", "fuel_system"))  # exclude 'make'

# Build formula as a string
formula_str <- paste("price ~", paste(c(quant_vars, qual_vars), collapse = " + "))

```

## 7: Regression Model
#### We run the regression model to evaluate the relationship between the selected predictor variables and the response variable (Price), observer their significancy and check the validity of the model.

``` {r Regression model}
# Fit the linear model
reg_model <- lm(as.formula(formula_str), data = auto_data_clean)

# View summary of the model
summary(reg_model)

# Residual Standard Error (RSE) from the model summary
rse <- summary(reg_model)$sigma

# Mean of the actual response variable
mean_price <- mean(auto_data_clean$price, na.rm = TRUE)

# Coefficient of Variation (as a percentage)
cv <- (rse / mean_price) * 100

# Print the result
cat("Coefficient of Variation (CV):", round(cv, 2), "%\n")
```

## 8: Residual Analysis 
#### Graphically identify the limitations of the model conducted to provide recommendations accordingly. 

```{r Residual Analysis}
plot(reg_model, 1)
plot(reg_model, 2)
plot(reg_model, 3)
plot(reg_model, 4)
plot(reg_model, 5)
plot(reg_model, 6)
par(mfrow = c(2, 3)) 
plot(reg_model, 1:6)
```

## 9: Actual vs Predicted Prices - ScatterPlot
#### Validate the efficiency of the model with a scatterplot of the actual prices versus the predicted priced. 
``` {r Actual prices vs predicted prices scatterplot}
# Create the same dataset used to fit the model
model_data <- model.frame(reg_model)  

# Add predictions to this dataset
model_data$predicted_price <- predict(reg_model)

# Plot: Actual vs Predicted Prices
library(ggplot2)
ggplot(model_data, aes(x = predicted_price, y = price)) +
  geom_point(color = "steelblue", alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Actual vs Predicted Prices",
       x = "Predicted Price",
       y = "Actual Price") +
  theme_minimal()


```
